Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
more incremental

$$ (J^t J + \lambda  \mathbf{1}_n)  \delta  = A \delta =  - J^t r $$

@@

The Gauss-Newton or Marquardt eqations are a set of linear equations. Moreover, 
the coefficient matrix is non-negative definite and symmetric. Thus it permits of
both general and specialized methods for linear equations. Furthermore, one can also
set up the solution without forming a coefficient matrix $A$ 
by using several matrix decomposition methods. Thus there are many possible procedures.

- Gauss elimination with partial or complete pivoting: The choice of partial or
  complete pivoting changes greatly the program code. However, neither of these
  choices uses the underlying structure and non-negative definite nature of 
  the matrix $A$.

- Variants of Gauss elimination that build matrix decompositions: We can apply
  elimination to produce an LU decomposition of $A$, and this may offer some 
  computational advantages, though in the present case such approaches do not
  seem to be ideal.

- Gauss-Jordan inversion: Here we compute the inverse of $A$ and then work with
  this inverse. That is, we use

  $$\delta = - A^{-1} g$$

@@

- Choleski Decomposition and back substitution: There are a number of varieties
  of Choleski decomposition, but they all take advantage of the symmetric and
  non-negative definite structure of $A$. A particularly compact form was used
  in @Nash1987a, though it is debatable whether the saving in storage was not
  heavily outweighed by the complexity of the code and its documentation.

- Eigendecompositions of the sum of squares and cross-products (SSCP) 
  matrix: the eigenvalues of real, symmetric matrices
  such as $A$ can be computed in many ways and the eigendecomposition used to 
  solve the Gauss-Newton equations. The computations are much more costly than
  simple solution of the equations, but the knowledge of the eigenvalues allows
  us to avoid, or at least be aware of, conditioning issues.

@@

Here we proposed to solve using the matrix $J$ or an augmented form $K$
directly. There are a number of approaches that can be used, which we
will divide into two classes, QR and SVD methods.

@@

There are many ways of computing such decompostions:

- Householder reflections with or without pivoting options
- Givens plane rotations, similarly
- Gram-Schmidt orthogonalization of the columns of $K$, either directly (one
  column at a time orhogonalized to all previous columns) or in a modified
  fashion that proceeds row-wise. There are a number of detailed choices.

@@

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements $\sqrt {(\lambda)}$ times the square roots of the 
diagonal stabilization elements
we wish to use. Then $K^t K$ is the appropriate $A$ for the Marquardt version
of the Gauss-Newton.

As we explain below, we do, unfortunately, realize that Marquardt suggested a scaled 
stabiilzation where we want to add the diagonals of $J^t J$ normal equation approach.
This requires that we perform at least a partial computation of the SSCP matrix,
which is relatively costly. For this reason we plan ?? to explore some proxies
for these elements.

@@

Thus it could be helpful to separate the computation of residuals from that of the
Jacobian. However, the structure of R is such that it is much easier to have the Jacobian
as an **attribute** of the residuals object. Indeed, it is the "gradient" attribute for 
historical reasons. This is the setup for both the finite difference approximations of
`nlspkg::numericDeriv()` (the Jacobian routine in base R for nls() in file `./src/library/stats/R/nls.R`)
as well as the `rjfun()` generated by `nlsr::model2rjfun()`. On the other hand, 
the computation of Jacobian elements can share many calculations with the computation of the residuals. 
Evidence is needed as to which de-duplication efforts give worthwhile savings, and in what
situations, but is likely of importance only to users with extremely time consuming problems.

## Storage stucture

@@

The original @jncnm79 Algorithm 23 (Marquardt nonlinear least squares solution) computes
the SSCP matrix $J^t J$  and solves the Marquardt-Nash augmented normal equations
with the Cholesky approach. This was continued in the Second Edition @nlacatvn1060620 and
in @jnmws87. However, in the now defunct @jnnlmrt12 and successor @nlsr-manual, the choice
has been to use a QR decomposition of an augmented Jacobian. `nls()` uses a QR decomposition
of the Jacobian without augmentation (referred to as the "gradient", however).

The particular QR calculations in these packages and in R-base are quite well-hidden in internal
code, complicating comparisons of storage, complexity and performance. The present exposition
is an attempt to provide some documentation.

?? sort out $A$ and $C$

@@

which is called the Choleski decomposition. In practice, we need $A$ computationally
non-singular (not just theoretically so), but the Marquardt stabilization can be used
to ensure this.

>>>>

$$ (J^t J + \lambda  \mathbf{1}_n)  \delta  = C \delta =  - J^t r $$

@@

The Gauss-Newton or Marquardt eqations are a set of linear equations. Moreover, 
the coefficient matrix is non-negative definite and symmetric. Thus it permits of
both general and specialized solution methods for linear equations. Furthermore, one can 
often set up the solution to a linear least squares problem without forming a coefficient matrix $C$ 
by using several matrix decomposition methods. Thus there are many possible procedures.

- Gauss elimination with partial or complete pivoting: The choice of partial or
  complete pivoting changes greatly the program code. However, neither of these
  choices uses the underlying structure and non-negative definite nature of 
  the matrix $A = J^t J$ or stabilized versions $C$ thereof.

- Variants of Gauss elimination that build matrix decompositions: We can apply
  elimination to produce an LU decomposition of $C$, and this may offer some 
  computational advantages, though in the present case such approaches do not
  seem to be ideal.

- Gauss-Jordan inversion: Here we compute the inverse of $C$ and then work with
  this inverse. That is, we use

  $$\delta = - C^{-1} g$$

@@

- Cholesky Decomposition and back substitution: There are a number of varieties
  of Cholesky decomposition, but they all take advantage of the symmetric and
  non-negative definite structure of $C$. A particularly compact form was used
  in @Nash1987a, though it is debatable whether the saving in storage was not
  heavily outweighed by the complexity of the code and its documentation.

- Eigendecompositions of the sum of squares and cross-products (SSCP) 
  matrix: the eigenvalues of real, symmetric matrices
  such as $C$ can be computed in many ways and the eigendecomposition used to 
  solve the Gauss-Newton equations. The computations are much more costly than
  simple solution of the equations, but the knowledge of the eigenvalues allows
  us to avoid, or at least be aware of, conditioning issues.

@@

Here we propose to solve the linear least squares sub-problem 
using the matrix $J$ or an augmented form $K$
directly. There are a number of approaches that can be used, which we
will divide into two classes, QR and singular value decomposition (SVD) methods.

@@

There are many ways of computing such decompositions:

- Householder reflections with or without pivoting options;
- Givens plane rotations, again with pivoting variants;
- Gram-Schmidt orthogonalization of the columns of $K$, either directly (one
  column at a time orhogonalized to all previous columns) or in a modified
  fashion that proceeds row-wise. There are a number of detailed choices.

@@

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt-Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements that the square roots of  $\lambda (D + \phi 1_n)$. 
Then $K^t K$ is the appropriate $C$ for the Marquardt version
of the Gauss-Newton. Clearly it is equivalent to augment $J$ by **both** $\lambda D$
and $\lambda \phi 1_n$.

Unfortunately, if we decide to use the stabilization involving $D$, then 
we must perform at least a partial computation of the SSCP matrix,
which is relatively costly. For this reason, it may be preferred to use
the unscaled stabilization using $1_n$ only. It may then be important to
provide an explicit (and fixed) parameter scaling.

?? manual scaling -- is it part of nlsr??

@@

Thus it could be helpful to separate the computation of residuals from that of the
Jacobian. However, the structure of R is such that it is often much easier to have the Jacobian
as an **attribute** of the residuals object. Indeed, it is the "gradient" attribute for 
historical reasons. This is the setup for both the finite difference approximations to the 
Jacobian in 
`nlspkg::numericDeriv()` which is the Jacobian approximation routine in base R for nls() in file
`./src/library/stats/R/nls.R`. It is also the structure of `rjfun()` generated by `nlsr::model2rjfun()`. 
On the other hand, 
the computation of Jacobian elements can share many calculations with the computation of the residuals. 
Evidence is needed as to which de-duplication efforts give worthwhile savings, and in what
situations, but is likely of importance only to users with extremely time consuming problems.

## Storage of algorithm objects

@@

The original @jncnm79 Algorithm 23 (Marquardt nonlinear least squares solution) computes
the SSCP matrix $J^t J$  and solves the Marquardt-Nash augmented normal equations
with the Cholesky approach. This was continued in the Second Edition @nlacatvn1060620 and
in @jnmws87. Even storing the full $A$ matrix uses only $n^2$ numbers.

However, in the now defunct @jnnlmrt12 and successor @nlsr-manual, the choice
has been to use a QR decomposition of an augmented Jacobian. Similarly, `nls()` uses a QR decomposition
of the Jacobian (referred to as the "gradient") without augmentation. Clearly the $K$ matrix is
at least $m$ by $n$, which for most nonlinear modelling problems is much larger than the $C$ matrix.
The particular QR calculations in `nls()` and its packaged version are quite well-hidden in internal
code, complicating comparisons of storage, complexity and performance.

@@

which is called the Cholesky decomposition. In practice, we need $A$ computationally
non-singular (not just theoretically so), but the Marquardt stabilization can be used
to ensure this.
